Sequential rank agreement methods for comparison of lists

### 2015-07-13 latex/paper.Rnw
Tweaks
Answers/rewrites to address Thomas' questions

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, the rankings within a single
list are not independent since each rank must appear exactly once in
each list. Thus, we cannot simply assign the same number (e.g., the
mean of the unassigned ranks) to the censored items since that would
% @Claus: less variation of the ranks? or of the agreement?
result in less variation of the agreement, and it would artificially
introduce a (downward) bias of agreement for items that are censored
in multiple lists.

@@

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
% @Claus: maybe H_1 is not a good name as people would expect H_1 to be the complement of H_0
\begin{eqnarray*}
\tilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the uncensored case each experiment contains
$L$ lists of random permutations of the numbers $1,\dots,P$. For the
censored/incomplete case we first permute the numbers $1,\dots,P$ and
then censor list $l$ at $d_l$. The sequential rank agreement curve
from the the original lists can then be compared to the pointwise rank
agreements obtained under $H_0$.

@@

For the analysis of the Golub data described in Section \ref{sec:amfol},
Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\tilde H_0$ each based on
$400$ permutations. Not surprisingly, the sequential rank agreement
under $\tilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank high) similar predictors even if there are only spurious
associations. The two bottom panels in
% @Claus jeg forstaar ikke den den naste lange saetning. mener du full lists i stedet for full dataset?
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement is better than what would be expected for the full
dataset but that the censored data (the bottom-right plot) suggests
that there may be at most 1 or 2 ranked items that are towards the top
of the lists that yield a result better than what would be expected.

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{\pi_l(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Here $|A|$ is the cardinality of the set $A$.
Originally the overlap was only considered for pairwise comparisons
but formula \eqref{def:overlap} can be applied when $L>2$.
% @Claus: har du en reference hvor OA bliver brugt paa to lister?
The average overlap at depth $d$ is defined as the average of the
overlaps until depth $d$,
\begin{equation}
AO(d) = \sum_{i=1}^d \frac{O(d)}{d}.
\end{equation}
This definition directly ensures that further emphasis is put on items
in the top of the lists since their item specific overlap contributes
to the calculation of average overlap at higher depths.

There are three main drawbacks of the average overlap method: First it
is highly sensitive to single items in the top of one the lists.
% @Claus: jeg forstaar ikke den naeste saetning.
The average overlap profile may change dramatically depending on the
occurrence of similar items in the lists. Secondly, when the number of
lists, $L$, is high then it might be difficult to obtain a non-zero
overlap (and hence a non-zero average overlap) at the beginning of the
lists because the overlap will be zero until an item is present in all
$L$ lists. Finally, the average overlap has an interpretation in terms
of moving averages of percentages (of $L$ sets) which is somewhat less
intuitive than our proposed average rank distance.

@@

%% #includegraphics marginalVsModel.pdf

@@

%% #includegraphics thomas.pdf

%% #includegraphics simulationTest2.pdf

>>>>

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, the rankings within a single
list are not independent since each rank must appear exactly once in
each list. Thus, we cannot simply assign the same number (e.g., the
mean of the unassigned ranks) to the censored items since that would
result in less variation of the ranks and hence less variation of the
agreement, and it would artificially introduce a (downward) bias of
agreement for items that are censored in multiple lists.

@@

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\tilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the uncensored case each experiment contains
$L$ lists of random permutations of the numbers $1,\dots,P$. For the
censored/incomplete case we first permute the numbers $1,\dots,P$ and
then censor list $l$ at $d_l$. The sequential rank agreement curve
from the the original lists can then be compared to the pointwise rank
agreements obtained under $H_0$.

@@

For the analysis of the Golub data described in Section
\ref{sec:amfol}, Figure~\ref{fig:example1} shows the empirical
distributions of sequential rank agreement under $H_0$ and
$\tilde H_0$ each based on $400$ permutations. Not surprisingly, the
sequential rank agreement under $\tilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank high) similar predictors even if
there are only spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement (the black line) is better than what would be expected
by chance for data that contain no information since it lies below the
reference areas. However, the censored data also suggests that there
may be at most 1 or 2 ranked items towards the top of the lists that
yield a result better than what would be expected (the bottom-right
plot).

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{\pi_l(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Here $|A|$ is the cardinality of the set $A$.  Typically the overlap
is only considered for pairwise comparisons but formula
\eqref{def:overlap} can be applied even when $L>2$
\citep{Bar-Ilan2006,Boulesteix2009}.

The average overlap at depth $d$ is defined as the average of the
overlaps until depth $d$,
\begin{equation}
AO(d) = \sum_{i=1}^d \frac{O(d)}{d}.
\end{equation}
This definition directly ensures that further emphasis is put on items
in the top of the lists since their item specific overlap contributes
to the calculation of average overlap at higher depths.

There are three main drawbacks of the average overlap method: First it
is highly sensitive to single items in the top of one the lists.  The
average overlap profile may change dramatically depending on the
presence or absence of similar items in the beginning of the
lists. For example, if all lists have the same item at the top then
the average overlap attains its maximum at 1, but if just one of the
lists differ then the overage overlap starts at its minimum of
0. Secondly, when the number of lists, $L$, is high then it might be
difficult to obtain a non-zero overlap (and hence a non-zero average
overlap) at the beginning of the lists because the overlap will be
zero until an item is present in all $L$ lists. Finally, the average
overlap has an interpretation in terms of moving averages of
percentages (of $L$ sets) which is somewhat less intuitive than our
proposed average rank distance.

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/marginalVsModel.pdf}
\end{center}
 \caption{Marginal}
 \label{fig:app1}
\end{figure}

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/thomas.pdf}
\end{center}
 \caption{Thomas}
 \label{fig:app2}
\end{figure}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/simulationTest2.pdf}
\end{center}
 \caption{SimulationTest2}
 \label{fig:app3}
\end{figure}
